Estimating Neighborhood Rents with Scraped Data

### 2018-06-28 manuscript/manuscript.Rmd
updating to include PDF output version

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess and Sarah Chasins"
date: "June 5, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: "header.html"
      after_body: "../footer.html"
abstract: "Online classified ads are a key source of information about local rental housing markets. This research first assesses the agreement of census tract rent estimates based on scraped rental advertisements with existing data sources. The second component of the paper investivates temporal and spatial dynamics in one-bedroom rents among Seattle neighborhoods using spatial models. All models' posterior distribution are estimated with integrated nested Laplace approximations (INLA). Training data for models cover quarterly tract estimates for 2017 Q2 through 2018 Q1, while the test data are 2018 Q2 median values as observed through June 4th. The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $100 for the training data and $140 for the test data."
---

<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #4b2e83;
}
</style>

@@

Demand for rental housing units in Seattle, WA surged in the past decade, due to relative scarcity in available housing amidst a period of [historic population growth](https://www.seattletimes.com/seattle-news/data/114000-more-people-seattle-now-this-decades-fastest-growing-big-city-in-all-of-united-states/){target="_blank"}. Substantial appreciation for almost 2 years straight raised Seattle home prices to levels where [most households are no longer able to buy a home](https://www.seattletimes.com/business/real-estate/seattle-area-homebuyers-needed-11000-pay-raise-in-past-year-to-keep-up-with-rising-prices/?utm_source=The+Seattle+Times&utm_campaign=de86e2d581-Morning_Brief_11_28_2017&utm_medium=email&utm_term=0_5beb38b61e-de86e2d581-122931741){target="_blank"} short of "driving till they qualify". More and more households must therefore rent to live in Seattle, and [almost half](http://www.jchs.harvard.edu/ARH_2017_cost_burdens_by_metro){target="_blank"} of these renter households pay 30% of their incomes on housing costs, if not more. Given the universal need for housing, the individual dynamics related to rising housing costs have a role in shaping neighborhood processes of "demographic inversion" like urban gentrification and [suburbanizing poverty](https://www.brookings.edu/testimonies/the-changing-geography-of-us-poverty/){target="_blank"} too.

@@

Existing public data are essentially limited to the [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs/) and [American Housing Survey (AHS)](https://www.census.gov/programs-surveys/ahs.html){target="_blank"} following the cessation of long-form decennial censuses in 2000. These data theoretically have two sources of error relative to the "true"" current median rent for a market and neighborhood approximations like census tracts. Census sources of rental information cannot provide insight into any change in rental cost associated with newer developments becoming available (at least not until considerable time has already passed) given the temporal delay in data publication. This feature of Census data is salient given newer housing units' generally higher rents and greater-than-average vacancy, both factors which lead to underestimating the present levels of rents for available housing units when either the ACS or decennial census are used as a current indicator. For already-leased rental housing units that are sampled, the temporal lag generates measurement error such that data on a given building's current asking rents may be underestimated relative to current tenants' payment levels.

@@

[Helena](http://helena-lang.org){target="_blank"} is a high-level programming language aimed at helping end users automate web scraping as a method of data collection. Users write programs in an extension for Google Chrome by demonstrating the first row of data they want to collect using clicks, form interactions and highlighting values. The extension translates an initial demonstration into a starting script written in Helena, which a user can further modify or set to run on a regular schedule. Though Helena programs collect data roughly at the speed they are recorded at, users can run multiple programs in parallel to increase the volume of data collected. Data accumulate on a remote server (soon-to-be AWS) where users can later query either the full table of scraped data or a subset scraped within a specified number of hours.

@@

For the UW team's data collection, a Helena program runs 24-hours a day and scrapes listings from the Seattle Craigslist. It is important to note that web scraping does violate Craigslist's Terms of Service, and that we are using all scraped data exclusively for academic research. For a more thorough discussion of this issue, consult page 3 of [Boeing and Waddell (2016)](https://arxiv.org/ftp/arxiv/papers/1605/1605.05397.pdf){target="_blank"}.

@@

The following series of graphics summarize the listing data collected to date and compare rent estimates from these data to ACS data and Dupre+Scott. In an attempt to compare summaries about like rental units as much as possible, visualizations focus on one-bedroom units in all but one instance. One-bedroom units are tied with the combined share of two and three bedroom units for the [most common unit bedroom size](https://factfinder.census.gov/bkmk/table/1.0/en/ACS/16_1YR/S2504/1600000US5363000){target="_blank"} in Seattle according to 2016 ACS data. For Figure 2, data limitations require that the comparison is all housing units since ACS only recently (i.e. 2015-2016) started publishing bedroom-specific rent levels.

@@

![**Figure 1.** _Map of Seattle (left), Seattle Craigslist one-bedroom listing point and 2D-density map (right)._](../output/map_point_density.png)

@@

![**Figure 2.** _Heatmap of N Seattle Craigslist one-bedroom listings scraped per day time-series._](../output/n_heatmap.png)

@@

![**Figure 3.** _Time-series of median rent (all bedroom sizes) for Seattle, 1997-2018. Historical values adjusted with CPI; Craigslist estimates are quarterly, Dupre+Scott are semi-annual (March and September), and ACS are annual (i.e. respective 1-Year Estimates for each year period)._](../output/rent_lineplot.png)

@@

The focus of the data description now turns towards estimating median rents for specific neighborhoods. While zip codes, unofficial City neighborhoods and even arbitrary grid geographies can all be used to aggregate micro-data about rental housing, we compare census tract estimates because they roughly approximate neighborhood-size units and can be linked with a variety of other data in Seattle and more generally. The following links provide reference maps for Seattle Census tracts ([2000](https://www.seattle.gov/Documents/Departments/OPCD/Demographics/GeographicFilesandMaps/2000CensusSeattleTractMap.pdf){target="_blank"} and [2010](https://www.seattle.gov/Documents/Departments/OPCD/Demographics/GeographicFilesandMaps/2010CensusTractsandZipCodeBoundaries.pdf){target="_blank"})

@@

![**Figure 4.** _Chloropleth map of neighborhood 1-bedroom rents across Craigslist, Dupre+Scott and ACS 2012-2016 5-Year Estimates. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties, while ACS releases median gross (asking) rents for all one-bedroom renter-occupied housing units sampled._](../output/rent_map_across_1B.png)

@@

![**Figure 5.** _Scatter plot of neighborhood 1-bedroom rents across Craigslist and Dupre+Scott. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties. Purple dashed line indicates loess fit._](../output/rent_scatter_across_1B.png)

@@

An alternative explanation for the difference between measures comes in considering the units these properties are sampling from in a given neighborhood. The greater difference between the measures among lower-rent neighborhoods (with Craigslist estimating these as more expensive than Dupre+Scott) provides some evidence for this theory. These tracts are found towards the North and South edges of the city, and generally are more residential neighborhoods where 20+ unit apartment complexes are not the entire stock of rental unit properties. In these tracts, Craigslist has a non-zero probability of sampling from any larger building but also small houses, duplex/triplexes, mother-in-law suites and recently, [Accessory Dwelling Units (ADUs)](http://www.seattle.gov/DPD/Publications/CAM/cam116a.pdf){target="_blank"}.

@@

We assume that the log of the median one-bedroom rent for Seattle neighborhoods is normally distributed and compute the posterior distributions of each model's hyperparameters and parameters $\theta_{it}$ using [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf){target="_blank"} with the default, weakly-informative priors for all random-effects. We choose to log the dependent variable because all values are positive and this transformation improves model fit.

@@

The BYM Spatial with AR(1) is the [Besag-York-Mollie (1991)](https://link.springer.com/article/10.1007/BF00116466){target="_blank"} intrinsic conditional autoregressive (ICAR) spatial model with AR(1) and exchangeable hyperparameters for the panel's temporal structure (referred to as `bym` using shorthand). The random-effect $v_i$ is assumed to come from a normal distribution whose mean is the mean of the neighbors' random effects and variance proportional to the number of neighbors (formally, $v_i|v_j, j \in ne(i) ~ \mbox{Normal}(\bar{v_i}, \frac{\sigma^{2}_v}{m_j})$. Lastly, the BYM Spatiotemporal AR(1) (`spt`) adds an exchangeable random-effect ($\delta_i \sim Normal(0, \tau_\delta)$) for each period to allow for area-specific deviations in time trends.

@@

Visit the interactive results dashboard [here](http://hesscl.com/smooth-sea/dashboard/index.html){target="_blank"}.

@@

The following tables show in-sample and out-of-sample (i.e. 2018 Q2) accuracy for the 5 model specifications in terms of root mean squared-error (RMSE) and mean absolute error (MAE). Note how though the Non-Spatial models (i.e. `ns` and `nsar1`) achieve slightly better in-sample fits, both Spatial models are more accurate at predicting the current month's surface of rent values. It is important to note how the goal in the present case is to understand the data generating process, that is, the systematic variation in the data less the random component. So while we want to minimize model error as a way to judge if our models improve on each other, we also want to smooth over sampling variability in the Craigslist data that induces local deviations compared to the true median rent. The [information criteria for the models](http://hesscl.com/smooth-sea/dashboard/index.html#model-fit){target="_blank"} generally show that the spatial specifications are more likely models of the neighborhood rent data generating process.

@@

![**Figure 6.** _Lineplot of Spatial AR(1) model predicted values. Colors denote a tract's deviation from the average neighborhood 1B rent level on a quarter-by-quarter basis._](../output/bym_lineplot.png)

@@

![**Figure 7.** _Chloropleth maps of actual versus predicted values (BYM Spatial AR(1) model) for the Q1 2017 to Q2 2018 period._](../output/avp_map_series.png)

@@

This study highlights the importance of considering scraped web data for understanding the dynamics of neighborhood rental housing costs. Throughout this work we demonstrate how Craigslist rental housing listings scraped may be used to research the estimated rent for the Seattle market and its neighborhoods. Using Bayesian spatial smoothing models further exploit the scraped data's spatial and temporal richness, giving researchers an option for studying rental housing market dynamics on a more granular scale where time is not only considered in terms of trend, but also random shock and seasonality.

Finally, this paper has endeavored (perhaps a little too much...) to provide an example of using scraped rental housing listings for research, but also to model how Rmarkdown, Git and GitHub are transformative  for our disciplines. There are no free lunches with reproducibility: these open-source technologies demand that we learn how to incorporate them into our work. To the extent that this requires thinking about analysis beyond a user base of 1, there can be some growing pains. However, there are some positive spillovers beyond improving the _science_ of social science---including new platforms and media for sharing research findings.

@@

[Boeing, G. and Waddell, P., 2017. New insights into rental housing markets across the united states: web scraping and analyzing craigslist rental listings. Journal of Planning Education and Research, 37(4), pp.457-476.](http://journals.sagepub.com/doi/abs/10.1177/0739456X16664789){target="_blank"}

[Besag, J., York, J. and Mollie, A., 1991. Bayesian image restoration with two applications in spatial statistics (with discussion) Annals of the Institute of Statistical Mathematics. 43, pp.1-59.](https://link.springer.com/article/10.1007/BF00116466){target="_blank"}

[Blangiardo, M., Cameletti, M., Baio, G. and Rue, H., 2013. Spatial and spatio-temporal models with R-INLA. Spatial and spatio-temporal epidemiology, 4, pp.33-49.](https://www.sciencedirect.com/science/article/pii/S1877584512000846){target="_blank"}

>>>>

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess and Sarah Chasins"
date: "June 5, 2018"
output:  pdf_document
abstract: "Online classified ads are a key source of information about local rental housing markets. This research first assesses the agreement of census tract rent estimates based on scraped rental advertisements with existing data sources. The second component of the paper investivates temporal and spatial dynamics in one-bedroom rents among Seattle neighborhoods using spatial models. Posterior distributions of all model hyperparameters and parameters are estimated with integrated nested Laplace approximations (INLA). Training data for models cover quarterly tract estimates for 2017 Q2 through 2018 Q1, while the test data are 2018 Q2 median values as observed through June 4th. The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $100 for the training data and $140 for the test data."
---

@@

Demand for rental housing units in Seattle, WA surged in the past decade, due to relative scarcity in available housing amidst a period of [historic population growth](https://www.seattletimes.com/seattle-news/data/114000-more-people-seattle-now-this-decades-fastest-growing-big-city-in-all-of-united-states/). Substantial appreciation for almost 2 years straight raised Seattle home prices to levels where [most households are no longer able to buy a home](https://www.seattletimes.com/business/real-estate/seattle-area-homebuyers-needed-11000-pay-raise-in-past-year-to-keep-up-with-rising-prices/?utm_source=The+Seattle+Times&utm_campaign=de86e2d581-Morning_Brief_11_28_2017&utm_medium=email&utm_term=0_5beb38b61e-de86e2d581-122931741) short of "driving till they qualify". More and more households must therefore rent to live in Seattle, and [almost half](http://www.jchs.harvard.edu/ARH_2017_cost_burdens_by_metro) of these renter households pay 30% of their incomes on housing costs, if not more. Given the universal need for housing, the individual dynamics related to rising housing costs have a role in shaping neighborhood processes of "demographic inversion" like urban gentrification and [suburbanizing poverty](https://www.brookings.edu/testimonies/the-changing-geography-of-us-poverty/) too.

@@

Existing public data are essentially limited to the [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs/) and [American Housing Survey (AHS)](https://www.census.gov/programs-surveys/ahs.html) following the cessation of long-form decennial censuses in 2000. These data theoretically have two sources of error relative to the "true"" current median rent for a market and neighborhood approximations like census tracts. Census sources of rental information cannot provide insight into any change in rental cost associated with newer developments becoming available (at least not until considerable time has already passed) given the temporal delay in data publication. This feature of Census data is salient given newer housing units' generally higher rents and greater-than-average vacancy, both factors which lead to underestimating the present levels of rents for available housing units when either the ACS or decennial census are used as a current indicator. For already-leased rental housing units that are sampled, the temporal lag generates measurement error such that data on a given building's current asking rents may be underestimated relative to current tenants' payment levels.

@@

[Helena](http://helena-lang.org) is a high-level programming language aimed at helping end users automate web scraping as a method of data collection. Users write programs in an extension for Google Chrome by demonstrating the first row of data they want to collect using clicks, form interactions and highlighting values. The extension translates an initial demonstration into a starting script written in Helena, which a user can further modify or set to run on a regular schedule. Though Helena programs collect data roughly at the speed they are recorded at, users can run multiple programs in parallel to increase the volume of data collected. Data accumulate on a remote server (soon-to-be AWS) where users can later query either the full table of scraped data or a subset scraped within a specified number of hours.

@@

For the UW team's data collection, a Helena program runs 24-hours a day and scrapes listings from the Seattle Craigslist. It is important to note that web scraping does violate Craigslist's Terms of Service, and that we are using all scraped data exclusively for academic research. For a more thorough discussion of this issue, consult page 3 of [Boeing and Waddell (2016)](https://arxiv.org/ftp/arxiv/papers/1605/1605.05397.pdf).

@@

The following series of graphics summarize the listing data collected to date and compare rent estimates from these data to ACS data and Dupre+Scott. In an attempt to compare summaries about like rental units as much as possible, visualizations focus on one-bedroom units in all but one instance. One-bedroom units are tied with the combined share of two and three bedroom units for the [most common unit bedroom size](https://factfinder.census.gov/bkmk/table/1.0/en/ACS/16_1YR/S2504/1600000US5363000) in Seattle according to 2016 ACS data. For Figure 2, data limitations require that the comparison is all housing units since ACS only recently (i.e. 2015-2016) started publishing bedroom-specific rent levels.

@@

![Map of Seattle (left), Seattle Craigslist one-bedroom listing point and 2D-density map (right).](../output/map_point_density.png)

@@

![Heatmap of N Seattle Craigslist one-bedroom listings scraped per day time-series.](../output/n_heatmap.png)

@@

![Time-series of median rent (all bedroom sizes) for Seattle, 1997-2018. Historical values adjusted with CPI; Craigslist estimates are quarterly, Dupre+Scott are semi-annual (March and September), and ACS are annual (i.e. respective 1-Year Estimates for each year period).](../output/rent_lineplot.png)

@@

The focus of the data description now turns towards estimating median rents for specific neighborhoods. While zip codes, unofficial City neighborhoods and even arbitrary grid geographies can all be used to aggregate micro-data about rental housing, we compare census tract estimates because they roughly approximate neighborhood-size units and can be linked with a variety of other data in Seattle and more generally. The following links provide reference maps for Seattle Census tracts ([2000](https://www.seattle.gov/Documents/Departments/OPCD/Demographics/GeographicFilesandMaps/2000CensusSeattleTractMap.pdf) and [2010](https://www.seattle.gov/Documents/Departments/OPCD/Demographics/GeographicFilesandMaps/2010CensusTractsandZipCodeBoundaries.pdf))

@@

![Chloropleth map of neighborhood 1-bedroom rents across Craigslist, Dupre+Scott and ACS 2012-2016 5-Year Estimates. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties, while ACS releases median gross (asking) rents for all one-bedroom renter-occupied housing units sampled.](../output/rent_map_across_1B.png)

@@

![Scatter plot of neighborhood 1-bedroom rents across Craigslist and Dupre+Scott. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties. Purple dashed line indicates loess fit.](../output/rent_scatter_across_1B.png)

@@

An alternative explanation for the difference between measures comes in considering the units these properties are sampling from in a given neighborhood. The greater difference between the measures among lower-rent neighborhoods (with Craigslist estimating these as more expensive than Dupre+Scott) provides some evidence for this theory. These tracts are found towards the North and South edges of the city, and generally are more residential neighborhoods where 20+ unit apartment complexes are not the entire stock of rental unit properties. In these tracts, Craigslist has a non-zero probability of sampling from any larger building but also small houses, duplex/triplexes, mother-in-law suites and recently, [Accessory Dwelling Units (ADUs)](http://www.seattle.gov/DPD/Publications/CAM/cam116a.pdf).

@@

We assume that the log of the median one-bedroom rent for Seattle neighborhoods is normally distributed and compute the posterior distributions of each model's hyperparameters and parameters $\theta_{it}$ using [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) with the default, weakly-informative priors for all random-effects. We choose to log the dependent variable because all values are positive and this transformation improves model fit.

@@

The BYM Spatial with AR(1) includes a random-effect drawn from the [Besag-York-Mollie (1991)](https://link.springer.com/article/10.1007/BF00116466) intrinsic conditional autoregressive (ICAR) distribution, along with AR(1) and exchangeable hyperparameters for the panel's temporal structure (referred to as `bym` using shorthand). The random-effect $v_i$ is assumed to come from a normal distribution whose mean is the mean of the neighbors' random effects and variance proportional to the number of neighbors (formally, $v_i|v_j, j \in ne(i) ~ \mbox{Normal}(\bar{v_i}, \frac{\sigma^{2}_v}{m_j})$. Lastly, the BYM Spatiotemporal AR(1) (`spt`) adds an exchangeable random-effect ($\delta_i \sim Normal(0, \tau_\delta)$) for each period to allow for area-specific deviations in time trends.

@@

Visit the interactive results dashboard [here](http://hesscl.com/smooth-sea/dashboard/index.html).

@@

The following tables show in-sample and out-of-sample (i.e. 2018 Q2) accuracy for the 5 model specifications in terms of root mean squared-error (RMSE) and mean absolute error (MAE). Note how though the Non-Spatial models (i.e. `ns` and `nsar1`) achieve slightly better in-sample fits, both Spatial models are more accurate at predicting the current month's surface of rent values. It is important to note how the goal in the present case is to understand the data generating process, that is, the systematic variation in the data less the random component. So while we want to minimize model error as a way to judge if our models improve on each other, we also want to smooth over sampling variability in the Craigslist data that induces local deviations compared to the true median rent. The [information criteria for the models](http://hesscl.com/smooth-sea/dashboard/index.html#model-fit) generally show that the spatial specifications are more likely models of the neighborhood rent data generating process.

@@

![Lineplot of Spatial AR(1) model predicted values. Colors denote a tract's deviation from the average neighborhood 1B rent level on a quarter-by-quarter basis.](../output/bym_lineplot.png)

@@

![Chloropleth maps of actual versus predicted values (BYM Spatial AR(1) model) for the Q1 2017 to Q2 2018 period.](../output/avp_map_series.png)

@@

This study highlights the importance of considering scraped web data for research about the dynamics of neighborhood rental housing costs. Throughout this work we demonstrate how Craigslist rental housing listings may be used to research the estimated rent for the Seattle market and its neighborhoods. Using Bayesian spatial smoothing models further exploits the scraped data's spatial and temporal richness, giving researchers an option for studying rental housing market dynamics on a more granular scale where time is not only considered in terms of trend, but also random shock and seasonality.

Finally, this paper has endeavored to provide a model of how Rmarkdown, Git and GitHub can be transformative for social science disciplines. There are no free lunches with reproducibility: these open-source technologies demand that researchers learn how to incorporate them into our work. While there can be growing pains to this process, there are some positive spillovers beyond improving the _science_ of social science---including new platforms and media for sharing research findings.

@@

[Boeing, G. and Waddell, P., 2017. New insights into rental housing markets across the united states: web scraping and analyzing craigslist rental listings. Journal of Planning Education and Research, 37(4), pp.457-476.](http://journals.sagepub.com/doi/abs/10.1177/0739456X16664789)

[Besag, J., York, J. and Mollie, A., 1991. Bayesian image restoration with two applications in spatial statistics (with discussion). Annals of the Institute of Statistical Mathematics. 43, pp.1-59.](https://link.springer.com/article/10.1007/BF00116466)

[Blangiardo, M., Cameletti, M., Baio, G. and Rue, H., 2013. Spatial and spatio-temporal models with R-INLA. Spatial and spatio-temporal epidemiology, 4, pp.33-49.](https://www.sciencedirect.com/science/article/pii/S1877584512000846)
